Visual diagnostics for constrained optimisation with application to guided tours

### 2020-12-09 paper/paper.Rmd
edit on reference, plot layout & captions, acknowleedgements

---
title: "Visual diagnostics for constrained optimisation with application to guided tours"
blinded: 0
author:
  - name: H.Sherry Zhang
    affiliation: Department of Econometrics and Business Statistics, Monash University 
    email:  Huize.Zhang@monash.edu
abstract: >
  Projection pursuit searches for interesting low-dimensional views of high-dimensional data via the optimisation of an index function. The initial paper by Friedman & Tukey in 1974 stated that  "the  technique  used  for  maximising  the  projection index strongly influences both the statistical and the computational aspects of the procedure." However, while many projection pursuit indices have been proposed in the literature, less work has been done on the optimisation procedures. In this paper we introduce visual diagnostics for optimisation algorithms, in particular those available in the projection pursuit guided tour. These diagnostics and workflows can be applied to a broad class of optimisers, to assess their performance. An R package, ferrn, has been created to implement the diagnostics.
keywords:
- optimisation
- projection pursuit 
- guided tour
- visualisation
- diagnostics 
- R
preamble: >
  \usepackage{amssymb, amsmath, mathtools, dsfont, bbm, array, booktabs}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
  \usepackage{todonotes}
output:
  bookdown::pdf_book:
    base_format: rticles::rjournal_article
    fig_height: 3
editor_options: 
  chunk_output_type: console

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      root.dir = here::here(), fig.path = here::here("figs/"), 
                      fig.align = 'center', cache = TRUE)
```

@@

Visualisation is widely used in exploratory data analysis (REF). Presenting information in a graphical format often allows people to uncover information they would otherwise not be aware of and provides a more comprehensive understanding of the problem at hand. The work presented in this paper combines data visualization with optimization methods. It aims at bring visualization tools into optimization problems to assist users to better understand the performance of their optimization methods in practice. This motivates our work of creating plots to diagnose optimisation algorithms which in turn can be also used not only to understand optimization performance but also to understand and compare the behaviour of different algorithms.

The goal of continuous optimisation is to find the best solution within the space of all feasible solutions where typically the best solution is decided by an objective function. Broadly speaking optimization can be unconstrained or constrained (@kelley1999iterative). The unconstrained problem can be formulated as a minimization (or maximization)  problem such as 
$min_{x} f(x)$ where $f:\mathbb{R}^n \rightarrow \mathbb{R}$ is an objective function with  certain properties defined in an $L^p$ space. In this case, solutions rely on gradient descent or ascent methods. In the constrained optimization problem additional restrictions  are introduced via a set of functions that can be convex or non-convex: $g_i:\mathbb{R}^n \rightarrow \mathbb{R}$ for $i = 1, \ldots k$ and hence the problem can be written as
$min_{x} f(x)$ *subject to* $g_i(x) \leq 0$. Here methods such as Langrange multipliers and convex optimization methods including linear and quadratic programming can be used.
<!-- $f: S \rightarrow \mathbb{R}$, with $S \in \mathbb{R}^n$, in a reduced space given by the problem constraints on $f$.  -->

The motivation of this paper is the optimisation problem arising in the projection pursuit guided tour (@buja2005computational) which is an exploratory data analysis tool that is defined to detect *interesting structures* or features  in high-dimensional data through a set of lower-dimensional projections that cover the entire high dimensional space using interpolation methods called tours (cook2008grand). The target of the optimisation is to identify the most *interesting* low-dimensional views of the data given by a corresponding projection matrix. The most *interesting* structures are formally defined by a function of projections, called index function which is optimized to uncover the most reveling structures in a high dimensional space (@cook1993projection).

The optimization challenges encountered in the projection pursuit guided tour problem are common to those of optimization in general. Examples of those include the existence of multiple maxima (local and global), the trade off between computational burden and proximity to the maxima,  dealing with noisy objective functions that might be non smooth and non diferentiable (@jones1998efficient). Those  are not unique to this context and therefore the visualization tools and optimization methods presented in this paper can be easily apply to any other optimization problems.

@@

In gradient descent methods the goal is to find the minimum of an objective function while in gradient ascent methods the interest is in the maximum. The minimum is searched for by taken steps  proportional to the negative gradient or of the function at the evaluation point (analogously positive gradient for gradient ascent methods), that is $x_{k+1} = x_k -\alpha_kS_k$ where $\alpha_k$ is the step size or the learning rate (@fletcher2013practical).  The step size is the key in the success of the optimization problem as inadequate step sizes might lead to a failure in convergence. These methods are known to be relative slow and can have convergence issues when problems are poorly conditioned (@trefethen1997numerical). In order to solve some of those problems, line minimization or line search methods were proposed (@shi2004convergence) where the step size is chosen in such a way that convergence is ensured. In line search methods users are required to provide an initial estimate $x_{1}$, a search direction $S_k$ for each iteration where the objective is to find the value of $\alpha_k \in \mathbb{R}$ that minimizes $f(x_k + \alpha_kS_k)$ with respect to $\alpha_k$ (this step is called the line search). Then move on to the next point following $x_{k+1} = x_k + \alpha_kS_k$ and repeat the process until the desire convergence is reached.
Modern development of line search methods focuses on improving the computation of the searching direction $S_{k}$, and on approximations of the step size $\alpha_k$ together with reducing the computational storing
burden of the intermediate computations while catering for practical optimisation problems. In the next section two line search methods for the optimisation used in the projection pursuit guided tour will be presented.

Although gradient optimization methods are very popular, they rely on the existence of the objective function derivatives and on the complexity of the constraints. Because of that, the second big class of optimization problems look at derivative free methods where the emphasis is on finding, in most cases, a near optimal solution. Examples of those  include response surface methodology (REF), stochastic approximation (REF),
random search (REF) and  heuristic methods like tabu search and genetic algorithms (REF). Later, we will present an optimization algorithm within the random search class, namely simulated annealing for optimization with the guided tour. Moreover, we will introduce an algorithm to further improve optimization solutions both in gradient descent methods and in derive free optimization.

Several R implementations address optimization problems with both general purpose as well as tasks specifics solvers. The most prominent one within the general solvers is *optim* in (REF-stats package) which have functions for gradient based and derivative free methods. Other packages having similar functionalities include *nloptr* (REF), XX.

@@

Mathematical details of the geodesic interpolation can be found in @buja2005computational. Figure \@ref(fig:tour-path) shows a sketch of the tour path. The blue frames are produced by the projection pursuit optimisation algorithm, and the white frames interpolate between them. The tour method has been implemented in the `tourr` package in R, available on the Comprehensive R Archive Network at [https://cran.r-project.org/web/packages/tourr/](https://cran.r-project.org/web/packages/tourr/) [@wickham2011tourrpackage].

```{r tour-path, out.height="30%", out.width="60%", fig.cap="(ref:tour-path)"}
include_graphics(here::here("img/tour_path.png"))
```

(ref:tour-path) Each square (frame) represents the projected data with a corresponding basis. Blue frames are found by an optimisation algorithm iteratively whilst the white frames are constructed between two blue frames by geodesic interpolation.

@@

The optimisation problem in the tour context is stated as follows:
....
.Given a randomly generated starting basis $\mathbf{A}_1$, projection pursuit finds the final projection basis $\mathbf{A}_T$ that satisfies the following optimisation problem:

@@

where $I_d$ is the $d$-dimensional identity matrix and the constraint requires the projection bases $\mathbf{A}$ to be orthogonal matrices.

@@

where $T^{\prime} = T + k_{J}+ l_{J+1}$. Note that there is no output in iteration $J + 1$ since the fact that it is the last iteration means that the optimiser cannot find a better basis and the algorithm terminates. In this notation, the final basis found is $A_T$ with the highest index value $I_T$.

@@

The data structure constructed above meets the tidy data principle [@wickham2014tidy] that requires each observation forms a row and each variable forms a column. With tidy data structure, data wrangling and visualisation have been significantly simplified by well-developed packages such as dplyr[@dplyr] and ggplot2[@ggplot2].

@@

```{r toy-search, fig.cap="\\label{fig:toy-search} A comparison of search\\_better with different values of the parameter max\\_tries. A six-variable dataset boa6 is used with holes index on a 2D problem. max\\_tries = 500 allows the optimiser to find better basis with higher index value at iteration six."}
```

@@

```{r toy-interp,fig.cap = "(ref:toy-interp)"}
```

(ref:toy-interp) The resulting trace plot on the interpolated points has been plotted when using three different algorithms to optimise the index. The color represents the number of iteration. It can be observed that each algorithm differs in length in the optimisation and the curvature of the improvement for each algorithm also varies.

@@

```{r toy-pca, fig.cap = "(ref:toy-pca)"}
```

(ref:toy-pca) 1D projection on the 5-variable dataset `boa5` with two optimisers: `search_better` and `search_geodesic`. The yellow point corresponds to the theoretical best basis [0, 1, 0, 0, 0] with `V2` being the only non-normal variable in the dataset.  The underlying grey points are randomly generated on the 5D space and reduced to 2D via PCA along with all the search points presented. The enlarged color points that colored as the interpolation are the starting points of each algorithm. They are initially simulated with the same starting points but all the bases in `search_geodesic` has been flipped positive to ensure that bases with the same projected images but a sign difference are represented by the same point in the plot.

Apart from checking the progression of an optimiser, another interesting aspect is to visualise how the search looks like in its parameter space. Given the orthonormality constraint, the projection bases $\mathbf{A}_{p \times d}$ live on the surface of a $p \times d$ dimension sphere, where the dimension can easily go over two or three. While visualising the search paths on the original high dimensional sphere would require skills for the viewers to perceive rotation of geometry in higher dimensional space ($d > 3$), an easier alternative is to view the reduced space via some dimension reduction methods, i.e. principal component analysis. To better perceive the search path as an embedding of a hollow sphere, random points on the high dimensional sphere is generated using the package `geozoo` and PCA is conducted on both the bases and the points on the surface of the sphere.

Figure \@ref(fig:toy-pca) plots the first two principal components of two search paths, one using `search_better` and the other using `search_geodesic`. The search in the space reduced by PCA matches with the optimiser description before where the random sampling in `search_better` is broader, controlled by `alpha` parameter, which is default to 0.5 while the directional search in `search_geodesic`, controlled by `delta` with a default of 0.01, is so tiny that it can barely be seen.

@@

```{r toy-pca-animated, fig.cap = "(ref:toy-pca-animated)"}
```

(ref:toy-pca-animated) A selected number of frames from the animated PCA plot. With animation, it is easier to track the progression from the start to finish in each algorithm.

Animated plots can be informative in diagnostics, especially in the case of PCA plot when the starting and ending of the search is not clear. Figure \@ref(fig:toy-pca-animated) shows six frames of an animated version of Figure \@ref(fig:toy-pca) and this time, it shows that `search_better` finds the optimum quicker than `search_geodesic`.

@@

```{r toy-tour, fig.cap="(ref:toy-tour)"}
```

(ref:toy-tour) A selected number of frames from the tour animation for viewing the 5D space of all the projection bases. The second frame on the top row views the space from a direction that is close to the one in the PCA plot. The tour animation allows for a more holistic view of the full space in high dimensions from different angles.

While viewing the bases on the reduced space via PCA shed some lights on the space the optimisers have explored, the visualisation on the original $p \times d$ dimension enables a more holistic stereoscopic view of the search. To view a high dimensional ($d \ge 3$) object on a screen, an approach is to play the rotation of the object in animation and this can be done via a regular grand tour. Compared to the PCA plot, the animated rotation (tour) displayed in Figure \@ref(fig:toy-tour) gives a more well-rounded view of the search and one can view the curved region of the tour path from different angles, which may not be presented in the PCA plot. Also the grand tour animation encompasses the PCA projection since the rotation from PCA is just one angle that maximises the variance of the bases and the grand tour produces a sequence of angles that view the search from different directions. As an evidence, the last frame in Figure \@ref(fig:toy-tour) is a frame select from the tour animation that is close to the PCA angle and the projection looks similar to the one in Figure \@ref(fig:toy-pca).

@@

In section \ref{toy-interp}, an interpolation with increase-then-decrease pattern has been presented. This pattern is undesirable since the optimiser could have started the next iteration from the highest basis on the tour path, as annotated as the interpolated basis in the plot, but instead, it is forced to start from the target basis. This motivates the design of an interruption to check the index value on the tour path so that the interpolating bases is accepted only up to the one with the largest index value. After implementing this interruption, the search finds a higher final index value with fewer steps as shown in the right panel of Figure \@ref(fig:interruption).

```{r interruption, fig.asp=0.5, fig.cap = "(ref:interruption)"}
```

(ref:interruption) Two-D projection on `boa6` data with holes index optimised by `search_better`.  A comparison of the trace plot with and without interruption. When the interruption does not take place, the index value of the target basis can be smaller than the interpolated basis. The interruption forces the interpolation to finish at the highest interpolated basis on the tour path.

@@

```{r polish, fig.height= 2.5, fig.cap = "(ref:polish)"}
```

(ref:polish) Two-D projection on `boa6` data with holes index optimised by `search_geodesic`. The left panel shows the final projected data before polish and the right panel shows the one after. The separation of the clusters on the y axis becomes sharper after the polish.

@@

Figure \@ref(fig:polish) presents the final projections found before and after applying `search_polish` on `search_geodesic`. Polish search improves the index value from `r get_best(holes_2d_geo)$index_val %>% format(digits = 4)`  to `r get_best(holes_2d_geo_polish)$index_val %>% format(digits = 4)` with reduction of weights on the non-informative variables. In terms of the projected data as in Figure \@ref(fig:polish), polish works to sharpen the edges of each cluster.

@@

<!-- ```{r noisy-better-geo,fig.asp=0.5,fig.cap = "(ref:noisy-better-geo)"} -->
<!-- ``` -->

(ref:noisy-better-geo) One-D projection on `boa5` data with noisy index `norm_kol` optimised by `search_geodesic` and `search_better`. The grey dashed line represents the index value of the theoretical best basis. `search_geodesic` fails to optimise the noisy index while `search_better` has made reasonable improvements to reach the index value close to the one of the theoretical best basis.

@@

Figure \@ref(fig:noisy-better-geo) compares the tracing plot of two optimisers: `search_geodesic` and `search_better`. This time, the interpolated path is no longer smooth when using either algorithm and  `search_geodesic` fails to optimise this index with barely any improvement of the index value. On the other hand, `search_better` is doing relatively well. With the theoretical best basis [`r matrix(c(0, 1, 0, 0, 0))`] producing an index value of `r kol_1d_better %>% bind_theoretical(matrix = matrix(c(0, 1, 0, 0, 0), nrow = 5, ncol = 1), index = tourr::norm_kol(nrow(boa5)), raw_data = boa5) %>% tail(1) %>% pull(index_val)`, `search_better` finds the final basis [`r get_best(kol_1d_better)$basis %>% .[[1]] %>% format(digits = 3)`] with an index value of `r get_best(kol_1d_better)$index_val %>% format(digits = 4)`. A further polish step will give a marginal improvement of index value to `r get_best(kol_1d_better_polish)$index_val` with a basis of [`r get_best(kol_1d_better_polish)$basis %>% .[[1]] %>% format(digits = 3)`]. At this stage, the difference between the theoretical best and what has been found is likely due to simulation error since the best possible basis for a simulated data will be slightly off the theoretical best basis, which is derived based on the distributional assumptions in Equations \ref{eq:sim-norm} to \ref{eq:sim-x7}.

```{r kol-better, fig.height = 6, fig.cap="(ref:kol-better)"}
```

(ref:kol-better) One-D projection of `norm_kol` index on `boa6` data optimised by `search_better` with 20 randomly generated seeds. Each point represents a basis in the original 6D space, reduced to 2D by PCA. The grey points are random bases generated from 6-D hollow sphere and the two yellow points represent the local maximum corresponds to basis when `V2` is found and the global maximum where `V7` is found. The points produced by the search algorithm is colored by whether the global or local maximum is found with the interpolated bases highlighted as a path and the final basis as an amplified point.

```{r kol-random, fig.height = 6, fig.cap = "(ref:kol-random)"}
```

(ref:kol-random) One-D projection of `norm_kol` index on `boa6` data optimised by `search_better_random` with the same seeds. `search_better_random` has a probabilistic acceptance implementation that would also accept a basis with lower index value. This design allows the optimiser to jump out of the local maximum and hence more instances find the global maximum.

```{r kol-random-tuned, fig.height= 6, fig.cap = "(ref:kol-random-tuned)"}
```

(ref:kol-random-tuned) One-D projection of `norm_kol` index on `boa6` data optimised by `search_better_random` with an larger searching neighbourhood of 0.7. Further tuning of the parameter has been taken place to adjust the neighbourhood parameter `alpha` from 0.5 to 0.7.

The second experiment with the noisy index is to understand how the optimisers perform when a local maximum is present. The dataset used is `boa6` where `x2` and `x7` are informative. The two theoretical best bases are [`r matrix(c(0, 1, 0, 0, 0, 0), nrow = 6)`] and [`r matrix(c(0, 0, 1, 0, 0, 0), nrow = 6)`] with index value `r index <- tourr::norm_kol(nrow(boa6)); proj <- as.matrix(boa6) %*% matrix(c(0, 1, 0, 0, 0, 0), nrow = 6); index(proj)` and `r index <- tourr::norm_kol(nrow(boa6)); proj <- as.matrix(boa6) %*% matrix(c(0, 0, 1, 0, 0, 0), nrow = 6); index(proj)`, respectively. Hence, the global maximum happens when variable `x7` is found. Simulation is done with 20 randomly generated seeds on two optimisers: `search_better` (Figure \@ref(fig:kol-better)) and `search_better_random` (Figure \@ref(fig:kol-random) and Figure \@ref(fig:kol-random-tuned)). Compared to Figure \@ref(fig:kol-better) and \@ref(fig:kol-random), Figure \@ref(fig:kol-random-tuned) further increases the neighbourhood parameter `alpha` from 0.5 to 0.7 to enlarge the search space.  The data object is collected for each simulation and computation is done to reduce the bases from its original 6D to a 2D points by principal component analysis so as to view the search paths.

>>>>

---
title: "Visual diagnostics for constrained optimisation with application to guided tours"
author:
  - name: H.Sherry Zhang
    affiliation: Department of Econometrics and Business Statistics, Monash University 
    email:  Huize.Zhang@monash.edu
abstract: >
  Projection pursuit searches for interesting low-dimensional views of high-dimensional data via the optimisation of an index function. The initial paper by Friedman & Tukey in 1974 stated that  "the  technique  used  for  maximising  the  projection index strongly influences both the statistical and the computational aspects of the procedure." However, while many projection pursuit indices have been proposed in the literature, less work has been done on the optimisation procedures. In this paper we introduce visual diagnostics for optimisation algorithms, in particular those available in the projection pursuit guided tour. These diagnostics and workflows can be applied to a broad class of optimisers, to assess their performance. An R package, ferrn, has been created to implement the diagnostics.
preamble: >
  \usepackage{amssymb, amsmath, mathtools, dsfont, bbm, array, booktabs}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
output: rticles::rjournal_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE,
                      root.dir = here::here(), 
                      fig.path = here::here("figs/"), 
                      cache = TRUE, 
                      fig.align = "center",
                      fig.height = 3,
                      out.width = "100%")
```

@@

Visualisation is widely used in exploratory data analysis [@tukey1977exploratory]. Presenting information in a graphical format often allows people to uncover information they would otherwise not be aware of and provides a more comprehensive understanding of the problem at hand. The work presented in this paper combines data visualization with optimization methods. It aims at bring visualization tools into optimization problems to assist users to better understand the performance of their optimization methods in practice. This motivates our work of creating plots to diagnose optimisation algorithms which in turn can be also used not only to understand optimization performance but also to understand and compare the behaviour of different algorithms.

The goal of continuous optimisation is to find the best solution within the space of all feasible solutions where typically the best solution is decided by an objective function. Broadly speaking optimization can be unconstrained or constrained [@kelley1999iterative]. The unconstrained problem can be formulated as a minimization (or maximization)  problem such as 
$\min_{x} f(x)$ where $f:\mathbb{R}^n \rightarrow \mathbb{R}$ is an objective function with  certain properties defined in an $L^p$ space. In this case, solutions rely on gradient descent or ascent methods. In the constrained optimization problem additional restrictions  are introduced via a set of functions that can be convex or non-convex: $g_i:\mathbb{R}^n \rightarrow \mathbb{R}$ for $i = 1, \ldots k$ and hence the problem can be written as
$\min_{x} f(x)$ *subject to* $g_i(x) \leq 0$. Here methods such as Langrange multipliers and convex optimization methods including linear and quadratic programming can be used.
<!-- $f: S \rightarrow \mathbb{R}$, with $S \in \mathbb{R}^n$, in a reduced space given by the problem constraints on $f$.  -->

The motivation of this paper is the optimisation problem arising in the projection pursuit guided tour [@buja2005computational] which is an exploratory data analysis tool that is defined to detect *interesting structures* or features  in high-dimensional data through a set of lower-dimensional projections that cover the entire high dimensional space using interpolation methods called tours [@cook2008grand]. The target of the optimisation is to identify the most *interesting* low-dimensional views of the data given by a corresponding projection matrix. The most *interesting* structures are formally defined by a function of projections, called index function which is optimized to uncover the most reveling structures in a high dimensional space [@cook1993projection].

The optimization challenges encountered in the projection pursuit guided tour problem are common to those of optimization in general. Examples of those include the existence of multiple maxima (local and global), the trade off between computational burden and proximity to the maxima,  dealing with noisy objective functions that might be non smooth and non diferentiable [@jones1998efficient]. Those  are not unique to this context and therefore the visualization tools and optimization methods presented in this paper can be easily apply to any other optimization problems.

@@

In gradient descent methods the goal is to find the minimum of an objective function while in gradient ascent methods the interest is in the maximum. The minimum is searched for by taken steps  proportional to the negative gradient or of the function at the evaluation point (analogously positive gradient for gradient ascent methods), that is $x_{k+1} = x_k -\alpha_kS_k$ where $\alpha_k$ is the step size or the learning rate [@fletcher2013practical].  The step size is the key in the success of the optimization problem as inadequate step sizes might lead to a failure in convergence. These methods are known to be relative slow and can have convergence issues when problems are poorly conditioned [@trefethen1997numerical]. In order to solve some of those problems, line minimization or line search methods were proposed [@shi2004convergence] where the step size is chosen in such a way that convergence is ensured. In line search methods users are required to provide an initial estimate $x_{1}$, a search direction $S_k$ for each iteration where the objective is to find the value of $\alpha_k \in \mathbb{R}$ that minimizes $f(x_k + \alpha_kS_k)$ with respect to $\alpha_k$ (this step is called the line search). Then move on to the next point following $x_{k+1} = x_k + \alpha_kS_k$ and repeat the process until the desire convergence is reached.
Modern development of line search methods focuses on improving the computation of the searching direction $S_{k}$, and on approximations of the step size $\alpha_k$ together with reducing the computational storing
burden of the intermediate computations while catering for practical optimisation problems. In the next section two line search methods for the optimisation used in the projection pursuit guided tour will be presented.

Although gradient optimization methods are very popular, they rely on the existence of the objective function derivatives and on the complexity of the constraints. Because of that, the second big class of optimization problems look at derivative free methods where the emphasis is on finding, in most cases, a near optimal solution. Examples of those include response surface methodology [@box1951experimental], stochastic approximation [@robbins1951stochastic], random search [@fu2015handbook] and  heuristic methods [@sorensen2013metaheuristics]. Later, we will present an optimization algorithm within the random search class, namely simulated annealing for optimization with the guided tour. Moreover, we will introduce an algorithm to further improve optimization solutions both in gradient descent methods and in derive free optimization.

Several R implementations address optimization problems with both general purpose as well as tasks specifics solvers. The most prominent one within the general solvers is `optim()` in the package \CRANpkg{stats}[@stats] which have functions for gradient based and derivative free methods. Other packages having similar functionalities include \CRANpkg{nloptr}.

@@

Mathematical details of the geodesic interpolation can be found in @buja2005computational. Figure \ref{fig:tour-path} shows a sketch of the tour path. The blue frames are produced by the projection pursuit optimisation algorithm, and the white frames interpolate between them. The tour method has been implemented in the R package \CRANpkg{tourr} [@tourr], available on the Comprehensive R Archive Network at [https://cran.r-project.org/web/packages/tourr/](https://cran.r-project.org/web/packages/tourr/)

```{r tour-path, out.height="25%",out.width="50%",fig.cap="Each square (frame) represents the projected data with a corresponding basis. Blue frames are found by an optimisation algorithm iteratively whilst the white frames are constructed between two blue frames by geodesic interpolation."}
include_graphics(here::here("img/tour_path.png"))
```

@@

pa
The optimisation problem in the tour context is stated as follows:
....
.Given a randomly generated starting basis $\mathbf{A}_1$, projection pursuit finds the final projection basis $\mathbf{A}_T$ that satisfies the following optimisation problem:

@@

\noindent where $I_d$ is the $d$-dimensional identity matrix and the constraint requires the projection bases $\mathbf{A}$ to be orthogonal matrices.

@@

\noindent where $T^{\prime} = T + k_{J}+ l_{J+1}$. Note that there is no output in iteration $J + 1$ since the fact that it is the last iteration means that the optimiser cannot find a better basis and the algorithm terminates. In this notation, the final basis found is $A_T$ with the highest index value $I_T$.

@@

The data structure constructed above meets the tidy data principle [@wickham2014tidy] that requires each observation forms a row and each variable forms a column. With tidy data structure, data wrangling and visualisation have been significantly simplified by well-developed packages such as \CRANpkg{dplyr} [@dplyr] and \CRANpkg{ggplot2} [@ggplot2].

@@

```{r toy-search, fig.width = 5.5, fig.cap="A comparison of \\code{search\\_better} with different values of the parameter \\code{max\\_tries}. A six-variable dataset \\code{boa6} is used with holes index on a 2D problem. \\code{max\\_tries = 500} allows the optimiser to find better basis with higher index value at iteration six."}
```

@@

```{r toy-interp,fig.width = 5.5, fig.cap = "The resulting trace plot on the interpolated points has been plotted when using three different algorithms to optimise the index. The color represents the number of iteration. It can be observed that each algorithm differs in length in the optimisation and the curvature of the improvement for each algorithm also varies."}
```

@@

```{r toy-pca, fig.cap = "1D projection on the 5-variable dataset \\code{boa5} with two optimisers: \\code{search\\_better} and \\code{search\\_geodesic}. The yellow point corresponds to the theoretical best basis [0, 1, 0, 0, 0] with `V2` being the only non-normal variable in the dataset.  The underlying grey points are randomly generated on the 5D space and reduced to 2D via PCA along with all the search points presented. The enlarged color points that colored as the interpolation are the starting points of each algorithm. They are initially simulated with the same starting points but all the bases in \\code{search\\_geodesic} has been flipped positive to ensure that bases with the same projected images but a sign difference are represented by the same point in the plot."}
```

Apart from checking the progression of an optimiser, another interesting aspect is to visualise how the search looks like in its parameter space. Given the orthonormality constraint, the projection bases $\mathbf{A}_{p \times d}$ live on the surface of a $p \times d$ dimension sphere, where the dimension can easily go over two or three. While visualising the search paths on the original high dimensional sphere would require skills for the viewers to perceive rotation of geometry in higher dimensional space ($d > 3$), an easier alternative is to view the reduced space via some dimension reduction methods, i.e. principal component analysis. To better perceive the search path as an embedding of a hollow sphere, random points on the high dimensional sphere is generated using the package \CRANpkg{geozoo} [@geozoo] and PCA is conducted on both the bases and the points on the surface of the sphere.

Figure \ref{fig:toy-pca} plots the first two principal components of two search paths, one using `search_better` and the other using `search_geodesic`. The search in the space reduced by PCA matches with the optimiser description before where the random sampling in `search_better` is broader, controlled by `alpha` parameter, which is default to 0.5 while the directional search in `search_geodesic`, controlled by `delta` with a default of 0.01, is so tiny that it can barely be seen.

@@

```{r toy-pca-animated, fig.cap = "A selected number of frames from the animated PCA plot. With animation, it is easier to track the progression from the start to finish in each algorithm."}
```

Animated plots can be informative in diagnostics, especially in the case of PCA plot when the starting and ending of the search is not clear. Figure \ref{fig:toy-pca-animated} shows six frames of an animated version of Figure \ref{fig:toy-pca} and this time, it shows that `search_better` finds the optimum quicker than `search_geodesic`.

@@

```{r toy-tour, fig.cap="A selected number of frames from the tour animation for viewing the 5D space of all the projection bases. The second frame on the top row views the space from a direction that is close to the one in the PCA plot. The tour animation allows for a more holistic view of the full space in high dimensions from different angles."}
```

While viewing the bases on the reduced space via PCA shed some lights on the space the optimisers have explored, the visualisation on the original $p \times d$ dimension enables a more holistic stereoscopic view of the search. To view a high dimensional ($d \ge 3$) object on a screen, an approach is to play the rotation of the object in animation and this can be done via a regular grand tour. Compared to the PCA plot, the animated rotation (tour) displayed in Figure \ref{fig:toy-tour} gives a more well-rounded view of the search and one can view the curved region of the tour path from different angles, which may not be presented in the PCA plot. Also the grand tour animation encompasses the PCA projection since the rotation from PCA is just one angle that maximises the variance of the bases and the grand tour produces a sequence of angles that view the search from different directions. As an evidence, the last frame in Figure \ref{fig:toy-tour} is a frame select from the tour animation that is close to the PCA angle and the projection looks similar to the one in Figure \ref{fig:toy-pca}.

@@

In section \ref{toy-interp}, an interpolation with increase-then-decrease pattern has been presented. This pattern is undesirable since the optimiser could have started the next iteration from the highest basis on the tour path, as annotated as the interpolated basis in the plot, but instead, it is forced to start from the target basis. This motivates the design of an interruption to check the index value on the tour path so that the interpolating bases is accepted only up to the one with the largest index value. After implementing this interruption, the search finds a higher final index value with fewer steps as shown in the right panel of Figure \ref{fig:interruption}.

```{r interruption, fig.asp=0.5, fig.cap = "Two-D projection on \\code{boa6} data with holes index optimised by \\code{search\\_better}.  A comparison of the trace plot with and without interruption. When the interruption does not take place, the index value of the target basis can be smaller than the interpolated basis. The interruption forces the interpolation to finish at the highest interpolated basis on the tour path."}
```

@@

```{r polish, fig.height= 2.5, fig.cap = "Two-D projection on \\code{boa6} data with holes index optimised by \\code{search\\_geodesic}. The left panel shows the final projected data before polish and the right panel shows the one after. The separation of the clusters on the y axis becomes sharper after the polish. "}
```

@@

Figure \ref{fig:polish} presents the final projections found before and after applying `search_polish` on `search_geodesic`. Polish search improves the index value from `r get_best(holes_2d_geo)$index_val %>% format(digits = 4)`  to `r get_best(holes_2d_geo_polish)$index_val %>% format(digits = 4)` with reduction of weights on the non-informative variables. In terms of the projected data as in Figure \ref{fig:polish}, polish works to sharpen the edges of each cluster.

@@

<!-- ```{r noisy-better-geo,fig.asp=0.5,fig.cap = "One-D projection on `boa5` data with noisy index `norm_kol` optimised by `search_geodesic` and `search_better`. The grey dashed line represents the index value of the theoretical best basis. `search_geodesic` fails to optimise the noisy index while `search_better` has made reasonable improvements to reach the index value close to the one of the theoretical best basis."} -->
<!-- ``` -->

@@

Figure \ref{fig:noisy-better-geo} compares the tracing plot of two optimisers: `search_geodesic` and `search_better`. This time, the interpolated path is no longer smooth when using either algorithm and  `search_geodesic` fails to optimise this index with barely any improvement of the index value. On the other hand, `search_better` is doing relatively well. With the theoretical best basis [`r matrix(c(0, 1, 0, 0, 0))`] producing an index value of `r kol_1d_better %>% bind_theoretical(matrix = matrix(c(0, 1, 0, 0, 0), nrow = 5, ncol = 1), index = tourr::norm_kol(nrow(boa5)), raw_data = boa5) %>% tail(1) %>% pull(index_val)`, `search_better` finds the final basis [`r get_best(kol_1d_better)$basis %>% .[[1]] %>% format(digits = 3)`] with an index value of `r get_best(kol_1d_better)$index_val %>% format(digits = 4)`. A further polish step will give a marginal improvement of index value to `r get_best(kol_1d_better_polish)$index_val` with a basis of [`r get_best(kol_1d_better_polish)$basis %>% .[[1]] %>% format(digits = 3)`]. At this stage, the difference between the theoretical best and what has been found is likely due to simulation error since the best possible basis for a simulated data will be slightly off the theoretical best basis, which is derived based on the distributional assumptions in Equations \ref{eq:sim-norm} to \ref{eq:sim-x7}.

```{r kol-better, fig.height = 6, fig.cap="One-D projection of \\code{norm\\_kol} index on \\code{boa6} data optimised by \\code{search\\_better} with 20 randomly generated seeds. Each point represents a basis in the original 6D space, reduced to 2D by PCA. The grey points are random bases generated from 6-D hollow sphere and the two yellow points represent the local maximum corresponds to basis when `V2` is found and the global maximum where `V7` is found. The points produced by the search algorithm is colored by whether the global or local maximum is found with the interpolated bases highlighted as a path and the final basis as an amplified point."}
```

```{r kol-random, fig.height = 6, fig.cap = "One-D projection of \\code{norm\\_kol} index on \\code{boa6} data optimised by \\code{search\\_better\\_random} with the same seeds. \\code{search\\_better\\_random} has a probabilistic acceptance implementation that would also accept a basis with lower index value. This design allows the optimiser to jump out of the local maximum and hence more instances find the global maximum."}
```

```{r kol-random-tuned, fig.height= 6, fig.cap = "One-D projection of \\code{norm\\_kol} index on \\code{boa6} data optimised by \\code{search\\_better\\_random} with an larger searching neighbourhood of 0.7. Further tuning of the parameter has been taken place to adjust the neighbourhood parameter `alpha` from 0.5 to 0.7. "}
```

The second experiment with the noisy index is to understand how the optimisers perform when a local maximum is present. The dataset used is \\code{boa6} where `x2` and `x7` are informative. The two theoretical best bases are [`r matrix(c(0, 1, 0, 0, 0, 0), nrow = 6)`] and [`r matrix(c(0, 0, 1, 0, 0, 0), nrow = 6)`] with index value `r index <- tourr::norm_kol(nrow(boa6)); proj <- as.matrix(boa6) %*% matrix(c(0, 1, 0, 0, 0, 0), nrow = 6); index(proj)` and `r index <- tourr::norm_kol(nrow(boa6)); proj <- as.matrix(boa6) %*% matrix(c(0, 0, 1, 0, 0, 0), nrow = 6); index(proj)`, respectively. Hence, the global maximum happens when variable `x7` is found. Simulation is done with 20 randomly generated seeds on two optimisers: `search_better` (Figure \ref{fig:kol-better}) and `search_better_random` (Figure \ref{fig:kol-random} and Figure \ref{fig:kol-random-tuned}). Compared to Figure \ref{fig:kol-better} and \ref{fig:kol-random}, Figure \ref{fig:kol-random-tuned}further increases the neighbourhood parameter `alpha` from 0.5 to 0.7 to enlarge the search space.  The data object is collected for each simulation and computation is done to reduce the bases from its original 6D to a 2D points by principal component analysis so as to view the search paths.

In Figure \ref{fig:kol-better}, the starting point largely affects whether `search_better` will find the global maximum. The three seeds on row three: seed 9145, seed 2511, and seed 9209, and three on row four: seed 2888, seed 9334, and seed 9819 easily find `V7` because they are born in Rome. On the other end, the ones on the first two rows can only find `V2`, largely due to the fact that they do not even have chances to search near `V7`.
